\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-12 lang-design.tex
stuff

In an OO language, a subclass can override methods of a superclass.
Similar expressiveness is useful for metatheory mechanization, too.
%
For example, in a derived family, the programmer may want to override
the proof of a lemma with a different proof strategy than it was proved
in the base family.
%
Overriding is potentially incompatible with having equalities on late bound names.

@@

Coq provides tactics for proving injectivity and disjointness of constructors
(i.e., \lsti{injection} and \lsti{discriminate}).
The proof terms generated by the tactics involve exhaustively
matching on the constructors of an inductive type, so
they do not work for extensible inductive types (\ref{chg:extensible-v-exhaustive}).
%
In principle, the programmer could use \lsti{FInduction} to prove
injectivity and disjointness.
But that would be tedious, and the programmer would have to revisit
the induction proof every time the inductive type is extended.
%
Having to use \lsti{FInduction} is unsatisfying, also because
injectivity and disjointness ought to hold
regardless of whether inductive types are extensible.

@@

The tactics \lsti{finjection} and \lsti{fdiscriminate} are powered by
\emph{partial recursors} that our plugin automatically generates for all
inductive types defined with \lsti{FInductive}.

As analyzed earlier, ordinary recursors, such as "tm_rec", are impossible
within a family in which the name of the inductive type is late bound.
However,
a key observation is that extensible inductive types still admit a
weakened form of recursors where the motive is an \lsti{option} type.
For example, within family "STLC", the partial recursor for
"tm" has the following type:

@@

%\noindent
Importantly, the partial recursor is compatible with the late binding of "tm" in its type.
When "STLC_tm_prec" is inherited into family "STLCFix", all the previous four equalities
still hold, and a trivial, fifth equality is made available:

@@

Although partial recursors appear weaker than ordinary recursors,
there is power in restraint:
they offer a principled way to derive injectivity and disjointness of
constructors, while allowing extensibility.
Partial recursors allow for injective mappings from values of a late bound inductive
type, like "tm", to values of an ordinary inductive type, like \lsti{\nat},
the injectivity and disjointness of whose constructors are readily
available.
%
For example, in a proof state with the assumption \lsti{H : tm_var "x" = tm_abs "y" t},
running \lsti{fdiscriminate H} first uses \lsti{H} to obtain the equality

@@

>>>>

In an OO language, a subclass can override methods of a superclass.
Similar expressiveness is useful for metatheory mechanization, too.
%
For example, in a derived family, the programmer may want to override
the proof of a lemma, using a different proof strategy than it was proved
in the base family.
%
Overriding is potentially incompatible with having equalities on late bound names, however.

@@

Coq provides tactics for proving injectivity and disjointness of constructors
(i.e., \lsti{injection} and \lsti{discriminate}).
The proof terms generated by the tactics involve exhaustively
matching on the constructors of an inductive type, so
they do not work for extensible inductive types (\ref{chg:extensible-v-exhaustive}).
%
In principle, the programmer could use \lsti{FInduction} to prove
injectivity and disjointness.
But this workaround is unsatisfying:
it is tedious, it forces the programmer to revisit
the induction proofs every time an inductive type is extended,
and above all, why should a property like
\lsti{tm_var "x" = tm_abs "y" t -> False} have anything to do with "tm_fix"?

@@

We make the observation
that injectivity and disjointness of existing constructors ought to hold
regardless of future addition of constructors.
%
This insight motivates the design of \emph{partial recursors}, which
power the \lsti{finjection} and \lsti{fdiscriminate} tactics.
Partial recursors are automatically generated for all inductive types
defined with \lsti{FInductive}.

As analyzed earlier, ordinary recursors, such as "tm_rec", are impossible
within a family in which the name of the inductive type is late bound.
However,
a key observation is that extensible inductive types still admit a
weakened elimination principle where the motive is an \lsti{option} type.
For example, within family "STLC", the partial recursor for
"tm" has the following type:

@@

%\noindent
Importantly, unlike the standard "tm_rec" recursor, the partial recursor
"STLC_tm_prec" is compatible with the late binding of "tm" in its type.
When "STLC_tm_prec" is inherited into family "STLCFix", all the previous four equalities
still hold, and a trivial, fifth equality is made available:

@@

Although partial recursors appear weaker than ordinary recursors,
there is power in restraint.
Partial recursors offer a principled, uniform way to derive injectivity and disjointness of
constructors, while allowing extensibility:
they enable injective mappings from values of a late bound inductive
type, like "tm", to values of an ordinary inductive type, like \lsti{\nat},
the injectivity and disjointness of whose constructors are readily
available.
%
For example, in a proof state with the assumption \lsti{H : tm_var "x" = tm_abs "y" t},
running \lsti{fdiscriminate H} first uses \lsti{H} to obtain the equality

@@

Note that the proof term generated by \lsti{fdiscriminate H} in family "STLC"
is reusable by family "STLCFix", because the partial recursor and its
equalities are compatible with the late binding of "tm".
Also note that within family "STLCFix", a second partial recursor
(called "STLCFix_tm_prec") on "tm" is available;
it can be used to derive properties about "tm_fix",
such as \lsti{tm_var "x" = tm_fix "y" t -> False}.
